---
title: "nba-clustering"
author: "Angel Manuel Velasquez"
date: "2024-11-21"
output: html_notebook
---

# Clustering

## Load Packages and data
```{r echo = FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggcorrplot))

# Data Source: https://www.kaggle.com/datasets/owenrocchi/nba-advanced-stats-20022022
nba_stats_processed <- readRDS("nba_stats_processed.rds")
names <- readRDS("names.rds")
nba_stats_processed$name <- names
```

## Helper functionss

### elbow method

```{r echo = FALSE:w}
perform_elbow_method <- function(df){
  WCSS <- sapply(1:7, function(k) kmeans(df,k)$tot.withinss)
  
  ggplot() +
    geom_point(aes(x = 1:7, y = WCSS), cex = 4) +
    geom_line(aes(x = 1:7, y = WCSS)) +
    labs(x = "Number of clusters (k)", 
         y = "Within-cluster sum of squares (WCSS)") +
    theme(text = element_text(size = 20))
}
```

### Fit a k-means model and visualize results
```{r echo = FALSE}
fit_and_visualize_model <- function(df, trainDf, k){
  
  print("Fitting model")
  kmModel <- kmeans(trainDf, k)
  df$cluster <- as.factor(kmModel$cluster)
  kmCenters <- as.data.frame(kmModel$centers)
  print("done")
  
  # Perform PCA
  print("Doing PCA")
  pca_result <- prcomp(trainDf, scale. = TRUE)
  print("done")
  
  # Calculate variance explained
  variance_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2) * 100
  
  # Get the loadings (feature contributions) for PC1 and PC2
  pc1_loadings <- abs(pca_result$rotation[, 1])
  pc2_loadings <- abs(pca_result$rotation[, 2])
  
  # Sort loadings to find top contributing features
  top_pc1_features <- names(sort(pc1_loadings, decreasing = TRUE)[1:3])
  top_pc2_features <- names(sort(pc2_loadings, decreasing = TRUE)[1:3])
  
  # Create a dataframe with PCA coordinates and cluster information
  pca_df <- data.frame(
    PC1 = pca_result$x[, 1],
    PC2 = pca_result$x[, 2],
    Cluster = df$cluster,
    Name = df$name
  )
  
  # Plot using ggplot2
  pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
    geom_point(alpha = 0.7) +
    geom_text(aes(label = Name), 
              size = 2, 
              vjust = -1, 
              check_overlap = TRUE) +
    theme_minimal() +
    labs(
      title = sprintf(
        "K-Means Clusters Visualized with PCA\nPC1 (%.1f%% var) Top Features: %s\nPC2 (%.1f%% var) Top Features: %s", 
        variance_explained[1], 
        paste(top_pc1_features, collapse = ", "),
        variance_explained[2],
        paste(top_pc2_features, collapse = ", ")
      ),
      x = "First Principal Component",
      y = "Second Principal Component"
    ) +
    scale_color_brewer(palette = "Set1") +
    theme(plot.title = element_text(size = 10))
  
  # Calculate the proportion of each position within clusters
  cluster_position_summary <- df %>%
    group_by(cluster, Pos) %>%
    summarise(count = n()) %>%
    group_by(cluster) %>%
    mutate(proportion = count / sum(count) * 100)
  
  pos_plot <- ggplot(cluster_position_summary, aes(x = cluster, y = count, fill = Pos)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      title = "Position Count within Clusters",
      x = "Cluster",
      y = "Number of Players",
      fill = "Position"
    ) +
    theme_minimal()
  
  combined_plot <- grid.arrange(pca_plot, 
                                pos_plot, 
                                ncol = 2,  
                                widths = c(2, 2),  # Adjust relative width of plots
                                heights = 1,          # Equal height
                                padding = unit(150, "lines")  # Add padding between plots
  )
    
}
```

## Cluster on all stats

### Prep data
```{r echo = FALSE}
# remove Pos. and name always
nba_stats_final <- nba_stats_processed |> select(-Pos, -name)

# use elbow to find optimal amount of clusters
perform_elbow_method(nba_stats_final)

```

### fit after doing elbow method
```{r echo = FALSE}
fit_and_visualize_model(nba_stats_processed, nba_stats_final, 4)
```

## Cluster on only advanced offense stats

## Process Data & elbow plot
```{r echo = FALSE}
# remove Pos. 
nba_stats_final <- nba_stats_processed |> select(-Pos)

# First, define the feature columns using tidyselect helpers
feature_columns <- c( "ORB.", "USG.",  "TS.", "AST.",  "TOV.", "X3PAr")
# Then create the training data using these columns

nba_stats_final <- nba_stats_processed |> 
  select( all_of(feature_columns))

perform_elbow_method(nba_stats_final)

```

### fit after doing elbow method
```{r echo = FALSE}
fit_and_visualize_model(nba_stats_processed, nba_stats_final, 4)
```


## Cluster on only advanced defensive stats

### Process Data 
```{r echo = FALSE}
# remove Pos. 
nba_stats_final <- nba_stats_processed |> select(-Pos)

feature_columns <- c( "DRB.", "STL.", "BLK.", "DBPM")

nba_stats_final <- nba_stats_processed |> 
  select( all_of(feature_columns))

perform_elbow_method(nba_stats_final)

```

### fit after doing elbow method
```{r echo = FALSE}
fit_and_visualize_model(nba_stats_processed, nba_stats_final, 3)
```

## Cluster on only advanced stats

### Process Data 
```{r echo = FALSE}
# remove Pos. 
nba_stats_final <- nba_stats_processed |> select(-Pos, -name)

feature_columns <- c( "VORP", "WS", "MP", "PER", "BPM")
# Then create the training data using these columns
nba_stats_final <- nba_stats_processed |> 
  select( all_of(feature_columns))
  
perform_elbow_method(nba_stats_final)

```

### fit after doing elbow method
```{r echo = FALSE}
fit_and_visualize_model(nba_stats_processed, nba_stats_final, 3)

```

## Cluster with no aggergate stats

### Process Data 
```{r echo = FALSE}
# remove Pos. 
nba_stats_final <- nba_stats_processed |> select(-Pos, -name)

feature_columns <- c("TS.", "X3PAr", "FTr", "TRB.", "AST.", "STL.", "BLK.", "VORP")
# Then create the training data using these columns
nba_stats_final <- nba_stats_processed |> 
  select( all_of(feature_columns))
  
perform_elbow_method(nba_stats_final)

```

### fit after doing elbow method
```{r echo = FALSE}
fit_and_visualize_model(nba_stats_processed, nba_stats_final, 4)

```

## Cluster with no physical stats and VORP/WS

### Process Data 
```{r echo = FALSE}
# remove Pos. 
nba_stats_final <- nba_stats_processed |> select(-Pos, -name)

feature_columns <- c("total_inches", "weight", "WS", "VORP", "Age")
# Then create the training data using these columns
nba_stats_final <- nba_stats_processed |> 
  select( all_of(feature_columns))
  
perform_elbow_method(nba_stats_final)

```

### fit after doing elbow method
```{r echo = FALSE}
fit_and_visualize_model(nba_stats_processed, nba_stats_final, 3)

```